Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-17 paper/paper.tex
made christophers edits before submitting

\section{Background: Detecting Sharing}

@@

The goal of detecting sharing is to generate a graph structure (like the AST in
the previous section) that share's common subexpressions. So we're essentially
going to generate a Directed Acyclic Graph (DAG) instead of an AST. For example,
we can use the following DAG structure that explicitly references nodes by a
unique identifier

@@

However the method does not work for DSL functions returning multiple outputs
via tuples or container types like lists. If we're attempting to generate a more
specific graph structure like a hypergraph or bi-partite graph the need to do
this becomes more aparent. For example, in our own research we implemented a DSL
for an instruction set architecture that generates a bipartite graph of
instructions and the registers they act upon. Attempting to generate a graph
where an instruction outputs multiple registers will result in returning
multiple independent state monads.

We can more concisely illustrate this problem by adding the following
instruction to our DSL that attempts to return two separate nodes in the DAG at
once
\begin{minted}{haskell}
novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}

@@

data DAG = DAG { unTrie :: Trie (Node,NodeID)
               , maxID  :: NodeID
               } deriving Show
\end{minted}
This looks a bit different because the \mintinline{haskell}{BiMap}
was a bijective relation between nodes and node ids,
whereas the trie maps byte strings to pairs (node,node id).
To get the DAG expressed as a relation, project out the values of the trie.

@@

For safety purposes, we need to keep track of a table of these labels and their
corresponding ASTs, to make sure we don't use the same label for different ASTs.
\begin{minted}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagSubMap :: Map ByteString ByteString
               , dagMaxID :: Int
               } deriving Show

@@

When a substitution is made via \mintinline{haskell}{subT}, the
\mintinline{haskell}{unStringAST} field is replaced with the new label, and the
previous serialized AST is placed in \mintinline{haskell}{unSubT}. When a
\mintinline{haskell}{Graph} value is processed, the \mintinline{haskell}{unSubT}
field is checked for if it contains a label

@@

Even with explicit sharing via substitutions, our method contains a reasonable
amount of overhead in order to overcome the shortcomings of Kiselyov's method.
The \mintinline{haskell}{addChains} example altered for explicit sharing with
both methods presents a worst case scenario in terms of overhead comparison.
Kiselyov's method is able to fully utilize its explicit sharing and our method
requires many substitution lookups.

@@

For more interesting benchmarks, we need to consider a more sophisticated DSL.
We can define a DSL for an instruction set architecture like mentioned in
section \ref{limitexplicit}. It will have the following interface

@@

We used this language to implement approximations of vector
\mintinline{haskell}{cos} and \mintinline{haskell}{tan} (i.e., a loop body that
computes over an input and output array). The loop body was unrolled by a factor
of four making for a sufficiently large resulting DAG.

We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e.,
hash-consing) and are own (i.e., trie-consing with some explicit sharing). It
wasn't possible to use Kiselyov's explicit sharing method where meaningful so
our code was able to achieve greater performance, 154\% speedup, with only a
small increase in memory consumption (see Fig.~\ref{fig:hashvscons}).

@@

For \mintinline{haskell}{tan}, we were unable to generate a resulting code graph using Kiselyov's
method. Using explicit sharing via let constructs wasn't possible due to the
multiple outputs issue and without it the amount of computation due to redundant
hash-consing was simply too large to perform in a reasonable amount of time.

>>>>

\section{Background: Detecting Sharing} \label{sec:background}

@@

The goal of detecting sharing is to generate a graph structure (like the AST in
Section \ref{sec:background}) but with sharing of common subexpressions. So
we're going to generate a Directed Acyclic Graph (DAG) instead of an
AST. For example, we can use the following DAG structure that explicitly
references nodes by a unique identifier

@@

However the method does not work for DSL functions returning multiple outputs
via tuples or container types like lists. Generating a hypergraph or bipartite
graph structure that allows operation nodes like \mintinline{haskell}{add} to
have edges to multiple outputs is one way this problem will arise. For example,
in our own research we implemented a DSL for an instruction set architecture
that generates a bipartite graph of instructions and the registers they act
upon. Attempting to generate a graph where an instruction outputs multiple
values will result in returning multiple independent state monads.

We can concisely illustrate this problem by adding the following
instruction to our DSL that attempts to return two separate nodes in the DAG at
once
\begin{minted}{haskell}
novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}

@@

data DAG = DAG { unTrie :: Trie (Node,NodeID)
               , maxID  :: NodeID
               } deriving Show
\end{minted}
This looks a bit different because the \mintinline{haskell}{BiMap} was a
bijective relation between nodes and node ids, whereas the trie maps byte
strings to pairs (node,node id). The DAG is expressed as a relation, by
projecting out the values of the trie.

@@

For safety purposes, we need to keep track of a table of these labels and their
corresponding ASTs, to make sure we don't use the same label for different
serialized ASTs.
\begin{minted}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagSubMap :: Map ByteString ByteString
               , dagMaxID :: Int
               } deriving Show

@@

When a substitution is made via \mintinline{haskell}{subT}, the
\mintinline{haskell}{unStringAST} field is replaced with the new label, and the
previous serialized AST is placed in \mintinline{haskell}{unSubT}. When a
\mintinline{haskell}{Graph} value is processed, the \mintinline{haskell}{unSubT}
field is checked to see if it contains a label

@@

Even with explicit sharing via substitutions, our method contains a reasonable
amount of overhead in order to overcome the limitations of Kiselyov's method.
The \mintinline{haskell}{addChains} example altered for explicit sharing with
both methods presents a worst case scenario in terms of overhead comparison.
Kiselyov's method is able to fully utilize its explicit sharing and our method
requires many substitution lookups.

@@

For more interesting benchmarks, we need to consider a more sophisticated DSL.
We defined a DSL for an instruction set architecture as mentioned in
section \ref{limitexplicit}. It will has the following interface

@@

We used this language to implement approximations of vector
\mintinline{haskell}{cos} and \mintinline{haskell}{tan} (i.e., a loop body that
cos or tan over an input array). The loop body was unrolled by a factor
of four resuling in a large DAG.

We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e.,
hash-consing) and our own (i.e., trie-consing with some explicit sharing). It
wasn't possible to use Kiselyov's explicit sharing method in this case so
our code was able to achieve greater performance, 154\% speedup, with only a
small increase in memory consumption (see Fig.~\ref{fig:hashvscons}).

@@

For \mintinline{haskell}{tan}, we were unable to generate a resulting code graph using Kiselyov's
method. Using explicit sharing via let constructs wasn't possible due to the
multiple outputs issue and without it the amount of computation due to redundant
hash-consing was simply too large, and would not terminate.
